\bf Model Visualization Techniques for a Social Network Model

### 2018-10-02 paper-child.Rnw
revise appendix

In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten. In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less. This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test. Since the tests are Wald-type tests, this makes sense.

@@

<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above. There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=

@@

To create the average net-vis shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times. As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data. Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1. Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3. In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2. None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.

@@

<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data. Outdegree and reciprocity have the same inverse relationship for both data sets. For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.", fig.height = 3>>=
simu2$data <- "friends"
senateEsts$data <- "senate"
allDataEsts <- rbind(senateEsts, simu2)
allDataEsts$parameter <- as.factor(allDataEsts$parameter)
allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) +
  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
  facet_grid(data~parameter, scales = "free", labeller = "label_both") +
  ThemeNoNet +
  theme(legend.position = 'bottom')
@

The density plots in Figure~\ref{fig:compareModels} show the distribution of the parameter estimates for the objective function parameters in M1, M2, and M3 the two example data sets we used. The most notable difference is in the values of $\beta_1, \beta_2$. In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while $\beta_3$ does seem to have an effect with the friendship data.

@@

In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3. It might therefore be advisable to consider only models that either allow $\beta_1$, or $\beta_2$. Looking at the high correlation with $\alpha$, we might switch to a model without $\beta_1$.

@@

Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}. To make this visualization, we first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations. Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later. We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations. Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases. Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000. Finally, we visualize all this information together in Figure~\ref{fig:alledgests}. In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process. We also see, however, that the process struggles to focus in on the edges in the second wave of the data. Ideally, we would like to see more occurrences of the edges which appear in the second wave of data. But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network. This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.

@@

Much of the code used to create these plots has been packaged an posted online at https://github.com/sctyner/netvizinf and https://github.com/sctyner/geomnet (also on CRAN). Additional, non-packaged code can be found at https://github.com/sctyner/SAOM-removing-blindfold.

>>>>

In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten. In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less. This suggests that larger data sets tend to result generally in smaller $p$-values, which is consistent with the construction of Wald-type tests.

@@

<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three expected networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above. There is some difference between the three models, but overall, these three models struggle to capture the structure in the true second wave of data.', fig.height=5>>=

@@

To create the expected network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each realized edge in the simulations.  The resulting summary network has weighted edges representing the proportion of edge appearances in all 1,000 simulations of wave 2. As in Figure~\ref{fig:summnetM1}, edges only appear in the expected networks if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the  expected network from the three models we fit and the first and second waves of data. Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1 but that model M2, which included the transitive triplet parameter, results in more strongly connected components. The lack of fit is clear: none of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In model M2, however, the ties to node 7 appear much weaker than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also strongly connected as they are in wave 2 of data, suggesting that M2 may be the best model of the three.

@@

<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data. %Outdegree and reciprocity have the same inverse relationship for both data sets. For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.", fig.height = 3>>=
simu2$data <- "friends"
senateEsts$data <- "senate"
allDataEsts <- rbind(senateEsts, simu2)
allDataEsts$parameter <- as.factor(allDataEsts$parameter)
allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) +
  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
  facet_grid(data~parameter, scales = "free", labeller = "label_both") +
  ThemeNoNet +
  theme(legend.position = 'bottom')
@

The density plots in Figure~\ref{fig:compareModels} show the distribution of the parameter estimates for the objective function parameters in M1, M2, and M3 the two example data sets we used. The most notable difference is in the values of $\beta_1, \beta_2$. In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while the opposite is true for the friendship data.

@@

In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation between them greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3. It might therefore be advisable to consider only models that either allow $\beta_1$ or $\beta_2$ but not both. Looking at the high correlation with $\alpha$, we might switch to a model without $\beta_1$.

@@

@@

Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process in Figure ~\ref{fig:alledgests}. To make this visualization, we first assign each possible edge in the network an ID number so that we can keep track of it throughout all microsteps and all simlations. Then, we count up the number of times each edge appears in the network throughout the microstep process for each of the 1,000 simulations. We also count the number of times an edge occurs at each step number. Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the step number increases. Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge was in the network in step 1, step 2, etc. divided by 1,000. Finally, we visualize all this information together in Figure~\ref{fig:alledgests}. In this plot, we see that all possible edges appear at least once at some point in the microsteps of at least one simulated process. We also see, however, that the process struggles to focus in on the edges in the second wave of the data. Ideally, we would like to see more occurrences the edges in wave 2 that are not in wave 1. But, about half of the edges in wave two are in the bottom half of the figure, which means they do not appear as much as they would if the model was actually capturing the mechanisms of tie change in the network. More of the darkest areas of the figure should belong to ties in wave 2, but those are often the lightest. This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.

\noindent \textbf{Online supplemental material:} Much of the code used to create these plots has been packaged an posted online at https://github.com/sctyner/netvizinf and https://github.com/sctyner/geomnet (also on CRAN). Additional, non-packaged code can be found at https://github.com/sctyner/SAOM-removing-blindfold.
